---
title: "Data Analysis - Exercise Sheet 1"
author: "Carl Herrmann, Maïwen Caudron-Herger"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
editor_options: 
  markdown: 
    wrap: sentence
---

```{r knitr_init, echo=FALSE, cache=FALSE}
library(knitr)
library(rmdformats)
library(dplyr)
library(tidyr)
library(ggplot2)
## Global options
options(max.print=200)
opts_chunk$set(echo=TRUE,
	             cache=TRUE,
               prompt=FALSE,
               tidy=TRUE,
               comment=NA,
               message=FALSE,
               warning=FALSE)
opts_knit$set(width=150)
```

# 1. Objectives

In this practical session you will

-   learn to use RStudio
-   learn how to implement some of the ideas taught in the first lesson using the R programming language.

Throughout this course we will use a common **diabetes dataset** to practically implement the concepts taught during the lectures.

Our dataset has various clinical measurements of \> 400 individuals.
Clinical parameters like blood glucose levels, cholesterol levels, age, body size, weight, blood pressure etc have been measured.

# 2. Loading and understanding the data

We will load a dataaset using the `read.delim` command in R, using an URL:

```{r}
dat <- read.delim('https://www.dropbox.com/scl/fi/zqjdoi7naolmruyxrraxq/diabetes_2025.tsv?rlkey=txnps3ccr0vj47yvjmpecbeom&dl=1') # Load the dataset
head(dat, 10) # Look at the first 10 lines of the table
```

In your RStudio "Environment", you should see a variable with the name "dat".
If you click on it, it will open this variable, that is a data.frame.

**1.** What is the dimension of our dataset (i.e. how many rows/columns are there in our data)

```{r}
# Dimension
dim(dat)
```

```{r, results='hide'}
# Number of columns
ncol(dat)
# Number of rows
nrow(dat)
```

**2.** What are the column names of our dataset

```{r}
colnames(dat) # Similarly rownames() for rows
```

For more description on these values [look here](https://hbiostat.org/data/repo/cdiabetes)

**3.** How do we extract the minimum and maximum age of patients in our dataset

```{r, results='hide'}
min(dat$age)
max(dat$age)
range(dat$age)
```

> Can you find out the same for height and weight?

**4.** How does the overall summary of our entire dataset look like?

```{r rows.print=10}
summary(dat)
```

> Can you explain what you see?

Before we keep going, here is a quick reminder of how to access columns, rows or individual cells in a dataframe:

**If you're already familiar with it you can skip the next chunk**.

```{r eval = FALSE}

# Returning a specific column or line of a data structure
dat[1,] # Returns the first line
dat[,1] # Returns the first column
dat[1,1] # Returns only the first object of the first line

# In a dataframe, if the columns have names, you can access the column using
dat$gender

# Returning an interval of columns or lines of a data structure 
dat[1:3,] # Returns the first three lines
dat[,1:3] # Returns the first three columns
dat[1:3,1:3] # Returns the first three elements of the first three lines

# Returning a value or an interval of values in a vector
dat$age[1] # Returns the first value
dat$age[1:3] # Returns the first three values

```

In this course we will often use the packages **dplyr** (for working with tables) and **ggplot2** (for plots). They are loaded at the top of this document. If R tells you that a package is not installed, run `install.packages("dplyr")`, `install.packages("tidyr")` and `install.packages("ggplot2")` once in the Console.

With **dplyr**, the same things can be written in a more readable way. The first argument is always the data frame:

```{r eval = TRUE}
head(select(dat, age, gender))                 # choose columns by name
head(select(dat, -age))                        # remove a column
head(filter(dat, age > 60))                    # choose rows with a condition
head(filter(dat, gender == 'male', age > 60))  # both conditions must be true
```

# 3. Data cleanup

Very often the first thing one needs to do before any data science project is to clean up the raw data and transform it into a format that is readily understood and easy to use for all downstream analysis.
This process usually involves --

-   Removing empty value rows/columns
-   Removing unused or unnecessary rows/columns
-   Reordering the data matrix
-   Keeping columns uniformly numeric (age, weight etc) or string (names, places etc) or logical (TRUE/FALSE, 1/0)
-   Handling strange caveats which are data specific like replacing `,` or `.`, or `;` from numbers etc

Lets do some clean up of our own diabetes data

1.  We will remove the `bp.2s` and `bp.2d` columns as it has mostly missing values (see summary above)
2.  We will also remove the column `time.ppn` which will not be required in our analysis
3.  We will reorder the columns of the data such that all the qualitative and quantitative values are separated. Among the quantitative values we will keep related variables together

To perform this cleanup, we use the following functions:

-   `select()` (package dplyr) to choose, remove and reorder columns
-   `is.na()` to find missing values
-   `drop_na()` (package tidyr) to remove rows with missing values

IMPORTANT: Before we do any cleaning, we save the original data into a new variable, so that we can go back to it if needed (for example, if we have done a mistake)

```{r}
dat.original <- dat
```

Now, we remove some columns with `select()`. The first argument is the data frame, followed by the names of the columns we want to remove, each with a "-" sign in front:

```{r}
dat <- select(dat, -bp.2s, -bp.2d, -time.ppn)
```

We re-order the remaining columns with `select()` as well: we simply write the names of the columns in the order in which we want them. We put the categorical columns first, and numerical columns after:

```{r}
dat <- select(dat, gender, location, frame, age, height, weight, waist, hip, stab.glu, glyhb, chol, hdl, ratio, bp.1s, bp.1d)
```

Now lets look at our cleaned data:

```{r}
summary(dat)
```

The ordering and selection of columns looks right, however it seems that there are certain columns that have missing values (like `glyhb` column has 13 `NA` values).
Lets remove all rows with any missing value: We go through each row and sum up all missing values (NA values).

Remember, 1 row = 1 patient.

For example, the 8th row (Patient with the number 1015) has a missing value in the column "bp.1s".
For this patient and column, the function `is.na()` will return the value "TRUE".

```{r}
is.na(dat[8,"bp.1s"]) # row number 8, column with the name 'bp.1s'
```

If you sum up "TRUE" values, each counts like a 1.
For example, TRUE + TRUE = 2.
Therefore, it is possible to use the function `sum()` on the result of the function `is.na()` applied on each row of the dataframe "dat".

For the first row alone, this would be obtained like this:

```{r}
sum(is.na(dat[1,]))
# This is calculating the number of NAs in the first row of the table "dat".

# The result is 0 because there are no NA on this line.
# Check out the result here:
dat[1,]

# The function is.na() returns only FALSE for each column of the first line.
is.na(dat[1,])
```

Instead of checking each patient (= each row) one by one, we can use the function `drop_na()`, which removes all rows containing at least one missing value. We store the number of rows before and after, to see how many patients were removed:

```{r}
n.before <- nrow(dat)
dat <- drop_na(dat)
n.after <- nrow(dat)
n.before - n.after
```

> How many patients were removed because they were associated with missing values?

Now our cleaned data has no missing values, columns are cleanly ordered and each column is in the right format

```{r}
summary(dat)
```

> Can you identify which types of data (continuous, discrete etc) each column above represents and why?

# 4. Visualizing data distribution

In this section you will learn the essential functions to plot data in an intuitive and useful way. We use the package **ggplot2**. A plot is built in layers, which are added with a `+`:

1.  `ggplot(data, aes(x = ...))` says which data frame to use, and which column goes on the x-axis (and y-axis, colors, ...). `aes` stands for *aesthetics*.
2.  `geom_...()` says what to draw (bars, points, boxes, ...).
3.  `labs()` sets the axis labels and the title.

### Histograms

We can plot the column "stab.glu" as a histogram using `geom_histogram()`:

```{r eval = TRUE}
ggplot(dat, aes(x = stab.glu)) +
  geom_histogram(bins = 30) +
  labs(x = "Stabilized Glucose concentration in blood", title = "Glucose concentration")
```

Remember that you can type `help("geom_histogram")` in the Console to see more details about this function.

> Change the parameter `bins = 30` to `bins = 50` in the above lines of code and see what happens.
> Try different values for `bins` like `10, 20, 75, 100` and try to interpret the differences.

### Density plots

For density plots, we use `geom_density()`, which estimates the probability density function for a given variable.
The parameter `bw` sets how strongly the curve is smoothed:

```{r eval = TRUE}
ggplot(dat, aes(x = stab.glu)) +
  geom_density(bw = 20) +
  labs(x = "Stabilized Glucose concentration in blood", title = "Glucose concentration")
```

> Change the value of `bw` and see how the plot changes

### Boxplots

`geom_boxplot()` produces a boxplot for a given variable. If we put the variable on the x-axis, the box is drawn horizontally:

```{r eval = TRUE}
ggplot(dat, aes(x = stab.glu)) +
  geom_boxplot() +
  labs(x = "Stabilized Glucose concentration in blood")
```

> Can you explain all features of this graph, such as upper/lower whisker, 25% quantile, ...?

### QQ-plots

As discussed in the lecture, we can use **QQ-plots** to either (1) compare two distributions, or (2) compare a distribution with a theoretical distribution (typically the normal distribution).

We can for example compare the distribution of the blood pressure values to check if they are normally distributed

```{r eval=TRUE}
## Let's first make a histogram
ggplot(dat, aes(x = bp.1s)) +
  geom_histogram(bins = 10)

## Maybe with more bins?
ggplot(dat, aes(x = bp.1s)) +
  geom_histogram(bins = 20)
```

Now we can use `geom_qq()` to generate the **QQ-plot** of this distribution against the standard normal distribution. Here we use the aesthetic `sample` instead of `x`:

```{r eval = TRUE}
ggplot(dat, aes(sample = bp.1s)) +
  geom_qq()
```

Using the additional layer `geom_qq_line()`, we can add a straight line that goes through the first and third quartile:

```{r eval = TRUE}
ggplot(dat, aes(sample = bp.1s)) +
  geom_qq() +
  geom_qq_line()
```

> So, is the distribution normal??

Now let's compare the quantiles of the cholesterol values of men and women!

```{r eval = TRUE}
# We can use "filter()" to select the male and female patients

men <- filter(dat, gender == 'male')
women <- filter(dat, gender == 'female')

# Compute the quantiles (note the "na.rm" option to ignore missing NA values!)
q.men <- quantile(men$bp.1s, probs = seq(0, 1, by = 0.05), na.rm = TRUE)
q.women <- quantile(women$bp.1s, probs = seq(0, 1, by = 0.05), na.rm = TRUE)

# Put both sets of quantiles into one table
q.table <- data.frame(q.men, q.women)

# Now plot against each other!
ggplot(q.table, aes(x = q.men, y = q.women)) +
  geom_point()
```

# 5. Exercises

### Exercise 1: Data Exploration and Summary Statistics

The summary() function gives us a quick overview, but sometimes we need more specific information.

1.  Calculate the mean and standard deviation of the hdl (High Density Lipoprotein) values for the entire dataset using `mean()` and `sd()`.
2.  Create a new variable called `bmi` (Body Mass Index) in the dataset using the formula: BMI = weight (in pounds) / height\^2 (in inches) × 703. Add this as a new column to `dat` using `mutate()`, i.e. `dat <- mutate(dat, bmi = ...)`
3.  Use the `table()` function on the column `location` to count how many patients come from each location. Which location has the most patients?
4.  Calculate the median age separately for male and female patients. Are they similar?
    *Hint: select the patients of one gender with `filter()` and compute the `median()` of their age.*

### Exercise 2: Visualizing Relationships Between Variables

In this exercise, you'll explore relationships between different clinical measurements.

1.  Create a scatter plot of `weight` vs `height` using `ggplot()` and `geom_point()`.
    Do you see a relationship between these variables? Change the aspect of the plot using the `size` parameter in `geom_point()`, for example `geom_point(size = 3)`.

2.  Color the points in your scatter plot from question 1 by gender by adding `color = gender` inside `aes()`.
    Does the relationship between weight and height appear different for men and women?

3.  Use the `cor()` function to calculate the correlation coefficient between weight and height.
    How strong is this relationship?

4.  Create side-by-side boxplots comparing cholesterol levels (`chol`) between males and females using `geom_boxplot()`, with `gender` on the x-axis and `chol` on the y-axis.
    What differences do you observe?
    Try with other variables!



### Going further: Advanced Data Manipulation (for advanced students)

1.  Create a categorical variable called `age_group` that divides patients into three categories: "young" (age \< 40), "middle" (40 ≤ age \< 60), and "senior" (age ≥ 60).
    Add this as a new column to `dat`.
    *Hint: use the `cut()` function inside `mutate()` and check how `cut()` works in the help*.

2.  Calculate the mean cholesterol level for each combination of `gender` and `age_group`.
    Store these results in a matrix or data frame.
    *Hint: use `group_by()` with `gender` and `age_group`, followed by `summarise()` with `mean()`*.

3.  Create a function called `standardize()` that takes a numeric vector as input and returns a z-transformed version (subtract mean, divide by standard deviation).
    Apply this function to all numerical columns in `dat` to create a standardized version of the dataset called `dat.scaled`.
    *Hint: you can use `apply()` with `MARGIN = 2` for columns, but remember to skip non-numeric columns*.

4.  Using your standardized dataset, identify which patient has the most extreme (highest or lowest) standardized BMI value.
    Print out all their clinical information.
    What makes this patient unusual?