---
title: "Exercise Sheet 06: hypothesis testing - solutions"
author: "Carl Herrmann, Maïwen Caudron-Herger"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  markdown: 
    wrap: sentence
---


```{r knitr_init, echo=FALSE, cache=FALSE}
library(knitr)
library(rmdformats)
library(dplyr)
library(ggplot2)
## Global options
options(max.print="120")
opts_chunk$set(echo=TRUE,
	             cache=TRUE,
               prompt=FALSE,
               tidy=TRUE,
               comment=NA,
               message=FALSE,
               warning=FALSE)
opts_knit$set(width=120)
opts_knit$set(root.dir  = "~/")
```

# 0. Recap of the previous sheet

In the last two exercises sheets you learnt about different major types of probability distributions like Normal, Binomial, Negative binomial, Student t etc.
You have also gained practical experience with the central limit theorem and the confidence intervals.

------------------------------------------------------------------------

# 1. Introduction and objectives

In this exercises sheet you will start to see how to *formulate hypotheses* and how to *test* them.
We want to learn how to use and interpret hypothesis tests.
We will mainly focus on tests on **mean values**.

------------------------------------------------------------------------

# 2. First statistical test

We will work again with the diabetes dataset that we used previously.

```{r}
dat = read.delim('https://tinyurl.com/y4fark9g')

# set the row names using the column id
rownames(dat) = dat$id
```

Check out the content of the dataset using the summary function

```{r}
summary(dat)
```

We could at first check how the weight values are distributed between males vs. females

```{r}
# Select the male and female patients
dat.men = filter(dat, gender == 'male')
dat.women = filter(dat, gender == 'female')

# Select the weight values accordingly
weight.men = dat.men$weight
weight.women = dat.women$weight

# Visualize the distributions as histograms (blue = men, orange = women)
p.weight = ggplot(dat, aes(x = weight, fill = gender)) +
  geom_histogram(aes(y = after_stat(density)), bins = 15, position = "identity", alpha = 0.6) +
  scale_fill_manual(values = c(male = "blue", female = "orange")) +
  labs(title = "Histogram of weight", x = "Weight")
p.weight
```

Now we could add the position of the means on the histogram

```{r}
mean.men <- mean(weight.men, na.rm = TRUE)
mean.women <- mean(weight.women, na.rm = TRUE)
#
p.weight +
  geom_vline(xintercept = mean.men, color = "blue", linetype = "dotted", linewidth = 1.5) +
  geom_vline(xintercept = mean.women, color = "orange", linetype = "dotted", linewidth = 1.5)
```

> Do you think that the mean of the male weight is lower than 180?
> Do you think that the mean of the female weight is different from the mean of the male weights?
> Do you think that the mean of the male weight is higher than that of the females?

------------------------------------------------------------------------

Now, we want to use a statistical test to check if,

-   males have a mean weight that is **significantly** lower than a specific value (one-sample, one-tailed test)
-   there is a **significant** difference in the mean of the weights between the two groups (two-sample, two-sided test)
-   male have a **significantly** higher mean weight than females (two-sample, one-sided test)

Note the use of the word **significant** in the previous statements!

This is exactly what **mean tests** such as the t-test (or the Wilcoxon test) are designed for!
We will perform here a **t-test**.

### One sample t-test

Using a one-sample t-test, we can check if values differ significantly from a target value.
For example, you could sample 10 chocolate bars, and test if they significantly differ from the expected weight of 100 g:

> Should we perform a one- or two-sided test?

```{r}
bars = c(103,103,97,102.5,100.5,103,101.3,99.5,101,104) # weights of 10 chocolate bars
chocbar.mean = 100 # expected weight
```

The function `t.test()` offers three **alternative** options: *two.sided*, *less* and *greater*.
Here, if we want to test whether the mean weight of the 10 chocolate bars is **different** from the expected weight of 100 g, we want to perform a *two sided* test and use the alternative *two.sided*.

It is essential to **clearly formulate the H0 and H1 hypothesis**.
There are two alternative but equivalent ways to do so.
Either:

\> H0: the expectation value of the random variable "Weight of a chocolate bar" is **equal** to 100 g.\

\> H1: the expectation value of the random variable "Weight of a chocolate bar" is **different** from 100 g.

or

\> H0: the mean weight of a chocolate bar is not significantly different from 100 g.\

\> H1: the mean weight of a chocolate bar is significantly different from 100 g.

Note the difference between these two formulations, and ask for help if you have questions about this!

```{r}
t.test(x = bars, mu = chocbar.mean, alternative = "two.sided")

# equivalent to:
# t.test(bars,mu = chocbar.mean)
# "alternative" is set to "two.sided" per default.

# x is a vector containing the data values
# mu indicates the true value of the mean (here: 100g)
```

> How would you interpret this result?
> Can you reject the H0 hypothesis?

Using alpha = 0.05, the H0 hypothesis can not be rejected as the p-value is 0.05244 (p-value \>= 0.05).
With alpha = 0.05, the mean weight of the chocolate bars is not significantly different from 100 g.

Using alpha = 0.1, the H0 hypothesis can be rejected (p-value \< 0.1).
With alpha = 0.1, the mean weight of the chocolate bars is significantly different from 100 g.

**BUT** ... It does not mean that alpha should be chosen with respect to the results of the t.test!!!

Before running a t.test, **formulate the null hypothesis H0 and the alternative hypothesis H1** and **decide about the alpha** value.
Remember, **alpha** represent the **false positive rate**: under the H0 hypothesis (test of two identical distributions), this is the proportion of tests that will detect a difference between the two groups (p-value \< alpha).

**Beware not to get confused between one-/two-sample tests, and one-/two-sided tests!**

Regarding the mean weight of the males, we would like to check whether males have a mean weight that is significantly **lower** than 180.
Here as well, we will perform a **one-sample test**, with mu = 180.
However, we will perform a **one-sided** test using the alternative option **less**.

The hypotheses can be formulated as:\
\> H0: the expectation value of the random variable "Weight of male patients" is **equal or greater** 180.\

\> H1: the expectation value of the random variable "Weight of male patients" is **less** than 180.

```{r}
t.test(weight.men, mu = 180, alternative = "less")
```

> How would you interpret the result with alpha = 0.05?
> Can you reject the H0 hypothesis?

### Two-sample t-test (2-sided)

Now, we will compare the mean of the weights between males and females.\

According to the previous histogram, females have a different mean weight as males.
This can be tested using a *two-sample and two.tailed* t.test.

```{r}
p.weight +
  geom_vline(xintercept = mean.men, color = "blue", linetype = "dotted", linewidth = 1.5) +
  geom_vline(xintercept = mean.women, color = "orange", linetype = "dotted", linewidth = 1.5)
```

The hypotheses can be formulated as:\
\> H0: the expectation value of the random variable "Weight of male patients" is **equal** to the expectation value of the random variable "Weight of female patients".\

\> H1: the expectation value of the random variable "Weight of male patients" is **different** from the expectation value of the random variable "Weight of female patients".

```{r}
t.test(weight.men, weight.women)
# remember, alternative = "two.sided" per default.
```

> How would you interpret the result with alpha = 0.05?
> Can you reject the H0 hypothesis?

### Two-sample t-test (1-sided)

Looking at the histogram, other observers could in principle see a difference between the mean values of the weights and formulate the following hypotheses:

\> H0: the expectation value of the random variable "Weight of male patients" is **equal or lower** to the expectation value of the random variable "Weight of female patients".\

\> H1: the expectation value of the random variable "Weight of male patients" is **higher** than the expectation value of the random variable "Weight of female patients".

```{r}
t.test(weight.men,weight.women, alternative = "greater")
```

> How would you interpret the result with alpha = 0.05?

> Can you explain why the p-value of the one-tailed t.test is lower than the p-value of the two-tailed t.test?
> What is the relation between these two values?

```{r}
# p-value of the two-sided t-test versus p-value of the one-sided t-test: 
t.test(weight.men,weight.women, alternative = "two.sided")$p.value # two-tailed
t.test(weight.men,weight.women, alternative = "greater")$p.value # one-tailed
# divide the one by the other and see ... 
```

This can be visualized using the t-distribution.
Here (see above result of the t.test), *t = 1.8453* and *df = 372.45*.

In the **one-tailed t.test**, the p-value is the area under the curve for t \> 1.8453 (alternative greater) **OR** for t \< -1.8453 (alternative less).

```{r}
# No need to understand this code - just look at the graph and shaded areas.
x = seq(-5,5,by=0.01)
y = dt(x,df=372.45)
t.dist = data.frame(x, y)
t.tail = filter(t.dist, x >= 1.84)   # the part of the curve with t > 1.84

ggplot(t.dist, aes(x = x, y = y)) +
  geom_line(linewidth = 1.5) +
  geom_area(data = t.tail, fill = "purple", alpha = 0.8) +
  geom_vline(xintercept = 1.8453, color = "blue", linetype = "dotted", linewidth = 1)
```

In the **two-tailed t.test**, the p-value is the area under the curve for t \> 1.8453 (alternative greater) **AND** for t \< -1.8453 (alternative less).
It is two times the p-value of the one-sided t.test!

```{r}
# No need to understand this code - just look at the graph and shaded areas.
x = seq(-5,5,by=0.01)
y = dt(x,df=372.45)
t.dist = data.frame(x, y)
t.tail.right = filter(t.dist, x >= 1.84)    # the part of the curve with t > 1.84
t.tail.left = filter(t.dist, x <= -1.84)    # the part of the curve with t < -1.84

ggplot(t.dist, aes(x = x, y = y)) +
  geom_line(linewidth = 1.5) +
  geom_area(data = t.tail.right, fill = "purple", alpha = 0.8) +
  geom_area(data = t.tail.left, fill = "purple", alpha = 0.8) +
  geom_vline(xintercept = c(-1.8453, 1.8453), color = "blue", linetype = "dotted", linewidth = 1)
```

IMPORTANTLY, a t-test can only be performed if the data is **normally distributed**!
We haven't check that yet for our dataset ... This is the topic of the section **going further** below.

------------------------------------------------------------------------

### Exercise 1

Consider the following graph, formulate the hypotheses H0 and H1 and perform a (one-sided) t-test.
Interpret the result using alpha = 0.05.

```{r}
ggplot(dat, aes(x = height, fill = gender)) +
  geom_histogram(aes(y = after_stat(density)), bins = 20, position = "identity", alpha = 0.6) +
  geom_vline(xintercept = mean(dat.men$height, na.rm = TRUE), color = "blue", linetype = "dotted", linewidth = 1.5) +
  geom_vline(xintercept = mean(dat.women$height, na.rm = TRUE), color = "orange", linetype = "dotted", linewidth = 1.5) +
  scale_fill_manual(values = c(male = "blue", female = "orange")) +
  labs(title = "Histogram of height", x = "Height")
```

> Solution: 

\> H0: the mean height value for the women group is **not significantly smaller** than the mean height for the men group.\
\> H1: the mean height value for the women group is **significantly smaller** than the mean height for the men group.

```{r}
# Two-sample t.test, one-tailed (alternative = "less")
t.test(dat.women$height, dat.men$height, alternative = "less")
```

With alpha = 0.05, the H0 hypothesis can be rejected (p-value \< alpha)

------------------------------------------------------------------------

### Exercise 2

1. Calculate the mean age of the men.

```{r}
### SOLUTION
age.men <- dat.men$age
mean(age.men)
```

2. Compare it to age = 50. Formulate the hypotheses H0 and H1 and perform a (one-sided) t-test. Interpret the result using alpha = 0.05.

> Solution:

\> H0: the expectation value of the random variable "age of male patients" is **greater or equal** to 50.\
\> H1: the expectation value of the random variable "age of male patients" is **less** than 50.

```{r}
# one-sample t-test, one-tailed (alternative = "less")
t.test(age.men, mu = 50, alternative = "less")
```

With alpha = 0.05, the H0 hypothesis cannot be rejected (p-value \>= alpha)


## ------------------------------------------------------------------------

# 3. Going further ... Checking the normality of the distribution

In principle, t-tests require that the data is approximately normally distributed.
If not, we can use **non-parametric** tests (see next lecture).

In order to check whether the data is normally distributed or not, it is possible to perform a **Shapiro-Wilk** normality test (see lecture).
This statistical test is implemented in R in the function `shapiro.test()`.

Let's have a look at the normality of the parameters in our diabetes dataset.

First, we perform a Shapiro-Wilk test on the *weight* after using a QQ-plot.

```{r}
# Check the normality of the distribution using a QQ-plot (Reminder)
ggplot(dat, aes(sample = weight)) +
  geom_qq() +
  geom_qq_line() +
  labs(title = "weight")
#
# Perform a Shapiro-Wilk test
shapiro.test(dat$weight)
# extract the p-value
shapiro.test(dat$weight)$p.value
# Reminder: for p-value >= 0.05, the data is normally distributed. For p-value < 0.05, the data is not normally distributed.
```

With a p-value \< 0.05, the data is NOT normally distributed!

We want to determine whether one of our numerical parameters is normally distributed:

```{r}
num.dat <- select(dat, chol, stab.glu, hdl, ratio, glyhb, age, height, weight, bp.1s, bp.1d, waist, hip) # create a dataframe containing only numeric columns
sapply(num.dat,shapiro.test) # apply the Shapiro-Wilk test on every column of "num.dat"
```

Apparently, none of the parameters is normally distributed.
For the purpose of the training, we will continue with "bp.1d" that does not look too bad ...

Here, with a QQ-plot:

```{r}
ggplot(dat, aes(sample = bp.1d)) +
  geom_qq() +
  geom_qq_line() +
  labs(title = "Blood pressure")
```

Now, we can perform a test on this parameters, using the male and female groups.
Considering the boxplots as seen below, we formulate the following hypotheses:

\> H0: the bp.1d mean value of the men group is not significantly different from the pb.1d mean value of the women group.\
\> H1: the bp.1d mean value of the men group is significantly different from the pb.1d mean value of the women group.

```{r}
ggplot(dat, aes(x = gender, y = bp.1d)) +
  geom_boxplot() +
  labs(title = "bp.1d values")
```

Accordingly, we perform a two-sample t.test (two-sided).
The result will be interpreted using alpha = 0.05.

```{r}
t.test(dat.men$bp.1d, dat.women$bp.1d, alternative = "two.sided")

# The observed t-statistics is:
t.obs = t.test(dat.men$bp.1d, dat.women$bp.1d, alternative = "two.sided")$statistic
```

With alpha = 0.05, the H0 hypothesis cannot be rejected!

------------------------------------------------------------------------

# 4. Extra insight: building the H0 distribution

The `t-test` is based on a theoretical distribution under the H0 hypothesis.
We can however try to build this H0 distribution empirically, using a sampling approach.

Let's compute the t-value for *bp.1d* for 2 **randomly chosen** groups within the cohort.
The rationale here: there is not reason why between 2 random groups, there should be a significant difference, right?

We create 2 **random** groups of patients, each one having the size of the *men* / *women* group

```{r}
# Sample random rows (according to the size of the male group)
i.randomMen = sample(1:nrow(dat),nrow(dat.men))
#
# Compute the t-test between the 2 random groups and get the t-value
t.h0=t.test(dat$bp.1d[i.randomMen],
            dat$bp.1d[-i.randomMen])$statistic
t.h0
```

> Repeat this several times to get various t-values over various random groups.

We can repeat this 10000 times over 10000 random splits of the cohorts:

```{r}
t.values = sapply(1:10000,function(i) {
  # random columns 
  i.randomMen = sample(1:nrow(dat),nrow(dat.men))
  #
  # get t-value between the 2 random groups
  t.test(dat$bp.1d[i.randomMen],
            dat$bp.1d[-i.randomMen])$statistic
})
#
# Plot the histogram
ggplot(data.frame(t.values), aes(x = t.values)) +
  geom_histogram(bins = 30)
```

This distribution approximates the $H_0$ distribution of the test statistics.

> What does the distribution look like??

We can compute the mean and standard deviation of the t.values and plot a qq-plot: 

```{r}
mean(t.values)
sd(t.values)
ggplot(data.frame(t.values), aes(sample = t.values)) +
  geom_qq() +
  geom_qq_line()
```

This is a **standard normal distribution** (actually, a t-distribution with degrees of freedom (df) equal to the number of patients -2)!\

> Do you remember why -2?

We can overlay the **empirical** $H_0$ distribution that we have just determined with the **theoretical** $H_0$ distribution, which is a t-distribution with `nrow(dat)-2` degrees of freedom.

```{r}
x = seq(-5,5,by=0.1)
y = dt(x,df=nrow(dat)-2)
dt.table = data.frame(x, y)
ggplot(data.frame(t.values), aes(x = t.values)) +
  geom_histogram(aes(y = after_stat(density)), bins = 30) +
  geom_line(data = dt.table, aes(x = x, y = y), color = "red", linewidth = 1.5)
```

> Pretty close, no?

## Computing the p-value

We can place the value of the observed t-statistics ("t.obs", see above) between the **real** men and women subgroups on this histogram:

```{r}
ggplot(data.frame(t.values), aes(x = t.values)) +
  geom_histogram(bins = 30) +
  geom_vline(xintercept = c(-t.obs, t.obs), color = "red", linetype = "dotted", linewidth = 1.2) +
  xlim(-5, 5)
```

Since we are performing a **2-sided** t-test here (remember the $H_0$ hypothesis formulation!), we have to check what the probability is to get **more extreme** t-values under $H_0$.

We can check in how many of these 10000 random splits we get a t-value that is more extreme than the observed, hence : lower than -abs(t.obs) or larger than abs(t.obs).
We then normalize this number to the 10000 random splits to obtain the p-value:

```{r}
p.value = sum(t.values < -abs(t.obs) | t.values > abs(t.obs))/length(t.values)
p.value
```

Instead of using the *empirical* $H_0$ distribution, we can use the *theoretical* distribution (t-distribution):

```{r}
## Degrees of freedom df
df = nrow(dat)-2
##
## this would be the one-sided p-value for the upper tail test (the probability of observing a t-value > t.obs)
pt(t.obs, df = df, lower.tail=FALSE)
##
## this is the two-sided p-value
2*pt(t.obs, df = df, lower.tail=FALSE)
```

Finally, we can compare these computed p-value with the p-value returned by the function `t.test`:

```{r}
t.test(dat.men$bp.1d, dat.women$bp.1d)
# You can extract the p-value using $p.value
t.test(dat.men$bp.1d, dat.women$bp.1d)$p.value
```

> It is pretty close, right?

One additional consideration: for large values of df (degree of freedom), the t-distribution is equivalent to the standard normal distribution N(0,1).

```{r}
x = seq(-4,4,by=0.1)
y = dt(x,df=nrow(dat)-2)  # built the t-distribution 
z = dnorm(x)  # built the standard normal distribution

# compare both distributions:
tn.table = data.frame(x = rep(x, times = 2),
                      density = c(y, z),
                      distribution = rep(c("t", "norm"), each = length(x)))
ggplot(tn.table, aes(x = x, y = density, color = distribution, linetype = distribution)) +
  geom_line(linewidth = 1.2) +
  scale_color_manual(values = c(t = "blue", norm = "red")) +
  labs(title = "t vs. norm for large df")
```

Appreciate the comparison below:

```{r}
# p-value based on the t-distribution
2*pt(t.obs, df = df, lower.tail=FALSE)
# p-value based on the standard normal distribution
2*pnorm(t.obs, lower.tail=FALSE)
```